---
title: 'Chicago Wildfire: PM2.5 Exposure Levels Across Sociodemographic Groups'
author: "Aaron Lilienfeld"
fig_width: 6
output:
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: 4
fig_height: 4
---

# Program background

* **Script name**: CW_Daily_PM25_Exposure_SES_ITS.RMD

* **Project name**: Chicago Wildfire 
   
* **Program description**: This program is meant to examine PM2.5 exposure among sociodemographic groups across the study period. 

* **Data**: 

* **Restrictions**:
   
* **Outline of program/analysis**:

* **Author**:	Aaron Lilienfeld         				

* **Date**:  

* **Revisions (by/date)**:

* **Input files**:

* **Output files**:
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = NA, fig.width = 10)
options(scipen = 999)
```

```{r load library}

library("tidyverse")
library("sf")
library("gt")
library("nlme")
library("AICcmodavg")
library("mgcv")

```

```{r load and process data, include=FALSE}

# Create neighborhood objects 
far_north_side <- toupper(c("Ohare", "O'HARE", "Norwood Park", "Edison Park", "Jefferson Park", "Forest Glen",
                            "North Park", "Albany Park", "Lincoln Square", "West Ridge", "Edgewater", "Rogers Park", "Uptown"))
northwest_side <- toupper(c("Dunning", "Montclare", "Belmont Cragin", "Hermosa", "Irving Park", "Portage Park"))
north_side <- toupper(c("Logan Square", "Avondale", "North Center", "Lake View", "Lincoln Park"))
west_side <- toupper(c("Austin", "Humboldt Park", "West Town", "West Garfield Park", "East Garfield Park", "Near West Side",
                       "North Lawndale", "South Lawndale", "Lower West Side"))
central <- toupper(c("Near North Side", "Loop", "Near South Side"))
southwest_side <- toupper(c("Garfield Ridge", "Clearing", "Archer Heights", "West Elsdon", "West Lawn",
                            "Chicago Lawn", "Gage Park", "Brighton Park", "McKinley Park", "New City",
                            "West Englewood", "Englewood"))
south_side <- toupper(c("Bridgeport", "Armour Square", "Douglas", "Fuller Park", "Grand Boulevard", "Greater Grand Crossing",
                        "Washington Park", "Kenwood", "Hyde Park", "Woodlawn", "South Shore", "Oakland"))
far_southwest_side <- toupper(c("Ashburn", "Auburn Gresham", "Beverly", "Mount Greenwood", "Morgan Park", "Washington Heights"))
far_southeast_side <- toupper(c("Chatham", "Roseland", "West Pullman", "Avalon Park", "Burnside",
                                "Pullman", "Riverdale", "Hegewisch", "South Deering", "South Chicago", "East Side", "Calumet Heights"))

# Load tract to community areas definition
trt_community_areas <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/Trt_Chicago_Community_Areas_2020.csv", 
                                         colClasses = c("GEOID20" = "character")) %>% 
  rename("geoid" = "GEOID20",
         "Community" = "COMMUNIT_1") %>% 
  mutate(Community = toupper(Community),
         Neighborhood = if_else(Community %in% far_north_side, "Far North Side",
                          if_else(Community %in% northwest_side, "Northwest Side", 
                                  if_else(Community %in% north_side, "North Side",
                                          if_else(Community %in% west_side, "West Side", 
                                                  if_else(Community %in% central, "Central", 
                                                          if_else(Community %in% southwest_side, "Southwest Side", 
                                                                  if_else(Community %in% south_side, "South Side",
                                                                          if_else(Community %in% far_southeast_side, "Far Southeast Side", 
                                                                                  if_else(Community %in% far_southwest_side, "Far Southwest Side", NA)))))))))) %>% 
  select(1,3,4) 

# Load tract shapefile 
trt_boundaries <- st_read("~/Documents/CEHI/Chicago_Wildfire/Data/tl_2020_17031_tract20/tl_2020_17031_tract20.shp") %>% 
  filter(ALAND20 != 0) %>% 
  rename("geoid" = "GEOID20") %>% 
  select(geoid) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  mutate(Community = str_to_title(Community),
         Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  st_set_crs(value = 26971) # NAD83 StatePlane Illinois East FIPS 1201

# Load Chicago community areas .shp file 
community_areas <- st_read("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/geo_export_933d2c08-f55b-413e-b529-b6b061fe738d.shp") %>% 
  st_set_crs(value = 26971) %>% 
  rename("Community" = "community") %>% 
  mutate(Neighborhood = if_else(Community %in% far_north_side, "Far North Side",
                          if_else(Community %in% northwest_side, "Northwest Side", 
                                  if_else(Community %in% north_side, "North Side",
                                          if_else(Community %in% west_side, "West Side", 
                                                  if_else(Community %in% central, "Central", 
                                                          if_else(Community %in% southwest_side, "Southwest Side", 
                                                                  if_else(Community %in% south_side, "South Side",
                                                                          if_else(Community %in% far_southeast_side, "Far Southeast Side", 
                                                                                  if_else(Community %in% far_southwest_side, "Far Southwest Side", NA)))))))))) %>% 
  select(Community, Neighborhood)

community_areas$Community[community_areas$Community == "OHARE"] <- "O'HARE"

# Chicago Health Atlas
community_areas_sociodemo <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas_CT.csv")[-c(1:4),] %>% 
  rename("geoid" = "GEOID",
         "Uninsured_rate" = "UNS_2018-2022",
         "COPD_rate" = "LNG_2021",
         "Disability_rate" = "DIS_2018-2022",
         "Unemployment_rate" = "UMP_2018-2022",
         "pctNHwht" = "PCT-W_2018-2022",
         "pctNHblk" = "PCT-B_2018-2022",
         "pctHisp" = "PCT-H_2018-2022",
         "HS_grad_rate" = "EDB_2018-2022",
         "Coll_grad_rate" =  "EDE_2018-2022",
         "Poverty_rate" = "POV_2018-2022",
         "MedHHinc" = "INC_2018-2022") %>% 
  select(3,13:15,10:12,8:9,5:7) %>% 
  mutate_at(.vars = c(2:12), as.numeric) %>% 
  mutate(across(2:12, round, 1)) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  select(1,13,14,2:12) %>% 
  mutate(Community = str_to_title(Community),
         Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Southwest Side", "Far Southwest Side", "Far Southeast Side")))

# Daily estimates files path
daily_pm25_trt_files <- list.files("~/Documents/CEHI/Chicago_Wildfire/Data/Output/PM25_Daily_Trt/",
                                   pattern = "2020",
                                   full.names = T)
# Dates of available data
days <- str_replace_all(seq.Date(as.Date("2023-05-01"), as.Date("2023-07-31"), by = "day"), "-", "")

# Save tracts listed in all files 
trt <- data.table::fread(daily_pm25_trt_files[1], colClasses = c("geoid" = "character"))[,1]

# Calculate daily avg
pm25_daily_avgs <- list()
for (i in 1:length(daily_pm25_trt_files)) {
  daily_pm25 <- data.table::fread(daily_pm25_trt_files[i], colClasses = c("geoid" = "character"))
  
  daily_pm25$PM25_Avg <- apply(daily_pm25[,2:25], FUN = mean, MARGIN = 1) # Calculate daily average based on hourly estimates
  daily_pm25 <- daily_pm25[,c(26)] # Subset to variable created 
  colnames(daily_pm25) <- paste0("PM25_Avg_", days[i]) # Change variable name according to date 
  pm25_daily_avgs[[i]] <- daily_pm25 
}

# Save into df 
pm25_daily_avgs <- cbind.data.frame(trt, pm25_daily_avgs)

# Setup data in long format 
pm25_long_fun <- function(df, values_to_label){
  pm25_daily_long <- df %>% 
    pivot_longer(cols = c(2:93), names_to = "Date", values_to = values_to_label) %>% 
    mutate(Date = as.Date(factor(Date, labels = seq.Date(as.Date("2023-05-01"), as.Date("2023-07-31"), by = "day"))))
  return(pm25_daily_long)
}

# Inner join with community areas 
pm25_daily_avgs_long <- pm25_long_fun(pm25_daily_avgs, "PM25_Avg") %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  mutate(Neighborhood = factor(Neighborhood, 
                               levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Southwest Side", "Far Southwest Side", "Far Southeast Side")))
```

# Assessing PM2.5 exposure levels pre-peak, peak, and post-peak wildfire smoke period across quartiles/quintiles of sociodemographic groups 

* As a reminder:
    + Pre-peak = May 1 - June 26 
    + Peak = June 27 - June 28 
    + Post-peak = June 29 - July 31

```{r process data for analysis - bin sociodemographic variables into quartiles/quintiles and join with daily average PM2.5 concentrations}

ses_quartiles <- community_areas_sociodemo
ses_quintiles <- community_areas_sociodemo

for (i in names(community_areas_sociodemo)[4:14]) {
  ses_quartiles[[i]] <- cut(ses_quartiles[[i]],
                            quantile(ses_quartiles[[i]], seq(0,1,0.25), na.rm = T),
                            include.lowest = T, 
                            labels = c("Q1", "Q2", "Q3", "Q4"))
  ses_quintiles[[i]] <- cut(ses_quintiles[[i]],
                            quantile(ses_quintiles[[i]], seq(0,1,0.2), na.rm = T),
                            include.lowest = T,
                            labels = c("Q1", "Q2", "Q3", "Q4", "Q5"))
}

# Join with daily PM2.5 
pm25_daily_avgs_quartiles <- pm25_daily_avgs_long %>%
  select(-Community, -Neighborhood) %>% 
  inner_join(ses_quartiles, by = "geoid") %>%
  mutate(Period = factor(if_else(Date <= as.Date("2023-06-26"), "Pre-peak",
                            if_else(Date >= as.Date("2023-06-29"), "Post-peak", "Peak")), levels = c("Pre-peak", "Peak", "Post-peak"))) 

pm25_daily_avgs_quintiles <- pm25_daily_avgs_long %>% 
  select(-Community, -Neighborhood) %>% 
  inner_join(ses_quintiles, by = "geoid") %>% 
  mutate(Period = factor(if_else(Date <= as.Date("2023-06-26"), "Pre-peak",
                            if_else(Date >= as.Date("2023-06-29"), "Post-peak", "Peak")), levels = c("Pre-peak", "Peak", "Post-peak")))

# Calculate ANOVA p-values 
anova_fun <- function(df, var, period){
  anova_formula <- as.formula(paste0("PM25_Avg ~ ", var))
  anova_pvalue <- round(summary(aov(anova_formula, data = df %>% filter(Period == period)))[[1]]$`Pr(>F)`[1],3)
  return(anova_pvalue)
}


pvals_quartiles_list <- list()
pvals_quintiles_list <- list()
for (i in names(pm25_daily_avgs_quartiles)[6:16]) {
 pvals_quartiles_list[[i]] <- c(anova_fun(pm25_daily_avgs_quartiles, var = i, period = "Pre-peak"), 
                                anova_fun(pm25_daily_avgs_quartiles, var = i, period = "Peak"),
                                anova_fun(pm25_daily_avgs_quartiles, var = i, period = "Post-peak"))
 
pvals_quintiles_list[[i]] <- c(anova_fun(pm25_daily_avgs_quintiles, var = i, period = "Pre-peak"), 
                                anova_fun(pm25_daily_avgs_quintiles, var = i, period = "Peak"),
                                anova_fun(pm25_daily_avgs_quintiles, var = i, period = "Post-peak"))
}

```

## Quartiles 

```{r setup table and output - quartiles}

summary_fun <- function(df, var){
  var <- enquo(var)
  df %>% 
    group_by(Period, !!var) %>% 
    summarise(PM25_Avg = paste0(round(mean(PM25_Avg), digits = 1), "(", round(sd(PM25_Avg), digits = 1), ")")) %>% 
    ungroup()
}

# Run function across variables 
# Save into list 
summary_dfs_quartiles_list <- list()
summary_dfs_quintiles_list <- list()
for (i in names(pm25_daily_avgs_quintiles)[6:16]) {
  summary_dfs_quartiles_list[[i]] <- summary_fun(pm25_daily_avgs_quartiles, var = get(i)) 
  summary_dfs_quintiles_list[[i]] <- summary_fun(pm25_daily_avgs_quintiles, var = get(i)) 
}

# Output table cleanly 
summary_dfs_quartiles_clean_list <- list()
summary_dfs_quintiles_clean_list <- list()
for (i in 1:length(summary_dfs_quartiles_list)) {
 summary_dfs_quartiles_clean_list[[i]] <- summary_dfs_quartiles_list[[i]] %>% 
    pivot_wider(names_from = `get(i)`, values_from = PM25_Avg) %>%
    mutate(`ANOVA p-val` = pvals_quartiles_list[[i]]) %>% 
    gt() %>% 
    tab_header(title = paste0("Average (SD) PM2.5 concentrations across quartiles of ", names(summary_dfs_quartiles_list)[i]))
 
  summary_dfs_quintiles_clean_list[[i]] <- summary_dfs_quintiles_list[[i]] %>% 
    pivot_wider(names_from = `get(i)`, values_from = PM25_Avg) %>%
    mutate(`ANOVA p-val` = pvals_quintiles_list[[i]]) %>% 
    gt() %>% 
    tab_header(title = paste0("Average (SD) PM2.5 concentrations across quintiles of ", names(summary_dfs_quintiles_list)[i]))
}

summary_dfs_quartiles_clean_list[[1]];summary_dfs_quartiles_clean_list[[2]];summary_dfs_quartiles_clean_list[[3]];summary_dfs_quartiles_clean_list[[4]];summary_dfs_quartiles_clean_list[[5]];summary_dfs_quartiles_clean_list[[6]];summary_dfs_quartiles_clean_list[[7]];summary_dfs_quartiles_clean_list[[8]];summary_dfs_quartiles_clean_list[[9]];summary_dfs_quartiles_clean_list[[10]];summary_dfs_quartiles_clean_list[[11]]

```

## Quintiles 

```{r setup table and output - quintiles}

summary_dfs_quintiles_clean_list[[1]];summary_dfs_quintiles_clean_list[[2]];summary_dfs_quintiles_clean_list[[3]];summary_dfs_quintiles_clean_list[[4]];summary_dfs_quintiles_clean_list[[5]];summary_dfs_quintiles_clean_list[[6]];summary_dfs_quintiles_clean_list[[7]];summary_dfs_quintiles_clean_list[[8]];summary_dfs_quintiles_clean_list[[9]];summary_dfs_quintiles_clean_list[[10]];summary_dfs_quintiles_clean_list[[11]]

```
