---
title: 'Chicago Wildfire: Main Analysis'
author: "Aaron Lilienfeld"
fig_width: 6
output:
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
  word_document:
    toc: yes
    toc_depth: 4
fig_height: 4
---

# Program background

* **Script name**: CW_Daily_PM25_Analysis_ITS.RMD

* **Project name**: Chicago Wildfire 
   
* **Program description**: 

* **Data**: 

* **Restrictions**:
   
* **Outline of program/analysis**:

* **Author**:	Aaron Lilienfeld         				

* **Date**:  

* **Revisions (by/date)**:

* **Input files**:

* **Output files**:
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = NA, fig.width = 10)
options(scipen = 999)
```

```{r load library}
# Load library
library("tidyverse")
library("sf")
library("kableExtra")
library("cowplot")
library("nlme")
library("table1")
library("AICcmodavg")
library("mgcv")
library("gratia")
library("zoo")

```

```{r load and process data, include=FALSE}

# Create neighborhood objects 
far_north_side <- toupper(c("Ohare", "O'HARE", "Norwood Park", "Edison Park", "Jefferson Park", "Forest Glen",
                            "North Park", "Albany Park", "Lincoln Square", "West Ridge", "Edgewater", "Rogers Park", "Uptown"))
northwest_side <- toupper(c("Dunning", "Montclare", "Belmont Cragin", "Hermosa", "Irving Park", "Portage Park"))
north_side <- toupper(c("Logan Square", "Avondale", "North Center", "Lake View", "Lincoln Park"))
west_side <- toupper(c("Austin", "Humboldt Park", "West Town", "West Garfield Park", "East Garfield Park", "Near West Side",
                       "North Lawndale", "South Lawndale", "Lower West Side"))
central <- toupper(c("Near North Side", "Loop", "Near South Side"))
southwest_side <- toupper(c("Garfield Ridge", "Clearing", "Archer Heights", "West Elsdon", "West Lawn",
                            "Chicago Lawn", "Gage Park", "Brighton Park", "McKinley Park", "New City",
                            "West Englewood", "Englewood"))
south_side <- toupper(c("Bridgeport", "Armour Square", "Douglas", "Fuller Park", "Grand Boulevard", "Greater Grand Crossing",
                        "Washington Park", "Kenwood", "Hyde Park", "Woodlawn", "South Shore", "Oakland"))
far_southwest_side <- toupper(c("Ashburn", "Auburn Gresham", "Beverly", "Mount Greenwood", "Morgan Park", "Washington Heights"))
far_southeast_side <- toupper(c("Chatham", "Roseland", "West Pullman", "Avalon Park", "Burnside",
                                "Pullman", "Riverdale", "Hegewisch", "South Deering", "South Chicago", "East Side", "Calumet Heights"))

# Load tract to community areas definition
trt_community_areas <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/Trt_Chicago_Community_Areas_2020.csv", 
                                         colClasses = c("GEOID20" = "character")) %>% 
  rename("geoid" = "GEOID20",
         "Community" = "COMMUNIT_1") %>% 
  mutate(Community = toupper(Community),
         Neighborhood = if_else(Community %in% far_north_side, "Far North Side",
                          if_else(Community %in% northwest_side, "Northwest Side", 
                                  if_else(Community %in% north_side, "North Side",
                                          if_else(Community %in% west_side, "West Side", 
                                                  if_else(Community %in% central, "Central", 
                                                          if_else(Community %in% southwest_side, "Southwest Side", 
                                                                  if_else(Community %in% south_side, "South Side",
                                                                          if_else(Community %in% far_southeast_side, "Far Southeast Side", 
                                                                                  if_else(Community %in% far_southwest_side, "Far Southwest Side", NA)))))))))) %>% 
  select(1,3,4) 

# Load tract shapefile 
trt_boundaries <- st_read("~/Documents/CEHI/Chicago_Wildfire/Data/tl_2020_17031_tract20/tl_2020_17031_tract20.shp") %>% 
  filter(ALAND20 != 0) %>% 
  rename("geoid" = "GEOID20") %>% 
  select(geoid) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  mutate(Community = str_to_title(Community),
         Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  st_set_crs(value = 26971) # NAD83 StatePlane Illinois East FIPS 1201
  
# Load Chicago community areas .shp file 
community_areas <- st_read("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/geo_export_933d2c08-f55b-413e-b529-b6b061fe738d.shp") %>% 
  st_set_crs(value = 26971) %>% 
  rename("Community" = "community") %>% 
  mutate(Neighborhood = if_else(Community %in% far_north_side, "Far North Side",
                          if_else(Community %in% northwest_side, "Northwest Side", 
                                  if_else(Community %in% north_side, "North Side",
                                          if_else(Community %in% west_side, "West Side", 
                                                  if_else(Community %in% central, "Central", 
                                                          if_else(Community %in% southwest_side, "Southwest Side", 
                                                                  if_else(Community %in% south_side, "South Side",
                                                                          if_else(Community %in% far_southeast_side, "Far Southeast Side", 
                                                                                  if_else(Community %in% far_southwest_side, "Far Southwest Side", NA)))))))))) %>% 
  select(Community, Neighborhood)

community_areas$Community[community_areas$Community == "OHARE"] <- "O'HARE"


# Chicago Health Atlas data 
# There are 5 separate data files to read in 
community_areas_sociodemo <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas_CT.csv")[-c(1:4),] %>% 
  rename("geoid" = "GEOID",
         "Uninsured_rate" = "UNS_2018-2022",
         "COPD_rate" = "LNG_2021",
         "Disability_rate" = "DIS_2018-2022",
         "Unemployment_rate" = "UMP_2018-2022",
         "pctNHwht" = "PCT-W_2018-2022",
         "pctNHblk" = "PCT-B_2018-2022",
         "pctHisp" = "PCT-H_2018-2022",
         "HS_grad_rate" = "EDB_2018-2022",
         "Coll_grad_rate" =  "EDE_2018-2022",
         "Poverty_rate" = "POV_2018-2022",
         "MedHHinc" = "INC_2018-2022") %>% 
  select(3,Population,13:15,10:12,8:9,5:7) %>% 
  mutate_at(.vars = c(2:13), as.numeric) %>% 
  mutate(across(3:13, round, 1)) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  select(1,14,15,2:13) %>% 
  mutate(Community = str_to_title(Community),
         Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Southwest Side", "Far Southwest Side", "Far Southeast Side")))

community_areas_sociodemo2 <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas_CT_v2.csv")[-c(1:4),] %>% 
  rename("geoid" = "GEOID",
         "Medicaid_rate" = "MCD_2019-2023",
         "Major_crimes" = "CZM_2020-2024",
         "Eviction_rate" = "EVR_2018",
         "Walkability_index" = "EKW_2024",
         "Rent_burdened_rate" = "RBU_2019-2023",
         "Ever_had_cancer_rate" = "CAN_2022",
         "Crowded_housing_rate" = "HTJ_2019-2023") %>% 
  select(geoid, Medicaid_rate, Major_crimes, Eviction_rate, Walkability_index, Rent_burdened_rate, Ever_had_cancer_rate, Crowded_housing_rate) %>% 
  mutate_at(.vars = c(2:8), as.numeric) %>% 
  mutate(across(2:8, round, 1)) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  select(1, 9, 10, 1:8)

community_areas_sociodemo3 <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas.csv")[-c(1:4),] %>% 
  rename("Community" = "Name",
         "Obesity_rate" = "HCSOBP_2023-2024",
         "Hypertension_rate" = "HCSHYTP_2023-2024", 
         "Diabetes_rate" = "HCSDIAP_2023-2024",
         "Asthma_rate" = "HCSATHP_2023-2024", 
         "Cancer_diagnosis_rate" = "CCR_2017-2021", 
         "Life_expectancy" = "VRLE_2023") %>% 
  select(Community, Obesity_rate, Hypertension_rate, Diabetes_rate, Asthma_rate, Cancer_diagnosis_rate, Life_expectancy) %>% 
  mutate_at(.vars = c(2:7), as.numeric) %>% 
  mutate(across(2:7, round, 1)) %>% 
  left_join((community_areas_sociodemo %>% distinct(Community, .keep_all = T) %>% select(Community, Neighborhood)), by = "Community") %>% 
  select(8,1, 2:7)

community_areas_sociodemo4 <- data.table::fread("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas_CT_v3.csv")[-c(1:4),] %>% 
  rename("geoid" = "GEOID",
         "pct_0_4" = "PCT-I_2019-2023",
         "pct_0_17" = "PCT-C_2019-2023", 
         "pct_65p" = "PCT-S_2019-2023") %>% 
  select(geoid, pct_0_4, pct_0_17, pct_65p) %>% 
  mutate_at(.vars = 2:4, as.numeric) %>% 
  mutate(across(2:4, round, 1)) %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  select(1,5,6,2:4)

community_areas_sociodemo5 <- readxl::read_excel("~/Documents/CEHI/Chicago_Wildfire/Data/Chicago_Community_Areas/CHA_Community_Areas_CT_v4.xlsx", sheet = 1) %>% 
  rename("Community" = "Name",
         "ADMR" = "VRADR_2020-2024") %>% 
  mutate(Community = toupper(Community)) %>% 
  inner_join(community_areas, by = "Community")

# Modify neighborhoods where NA appears due to spelling issues 
community_areas_sociodemo3[community_areas_sociodemo3$Community == "McKinley Park", "Neighborhood"] <- "Southwest Side"
community_areas_sociodemo3[community_areas_sociodemo3$Community == "O'Hare", "Neighborhood"] <- "Far North Side"

# Daily estimates files path
daily_pm25_trt_files <- list.files("~/Documents/CEHI/Chicago_Wildfire/Data/Output/PM25_Daily_Trt/",
                                   pattern = "2020",
                                   full.names = T)
# Dates of available data
days <- str_replace_all(seq.Date(as.Date("2023-05-01"), as.Date("2023-07-31"), by = "day"), "-", "")

# Save tracts listed in all files 
trt <- data.table::fread(daily_pm25_trt_files[1], colClasses = c("geoid" = "character"))[,1]

# Calculate daily avg
pm25_daily_avgs <- list()
for (i in 1:length(daily_pm25_trt_files)) {
  daily_pm25 <- data.table::fread(daily_pm25_trt_files[i], colClasses = c("geoid" = "character"))
  
  daily_pm25$PM25_Avg <- apply(daily_pm25[,2:25], FUN = mean, MARGIN = 1) # Calculate daily average based on hourly estimates
  daily_pm25 <- daily_pm25[,c(26)] # Subset to variable created 
  colnames(daily_pm25) <- paste0("PM25_Avg_", days[i]) # Change variable name according to date 
  pm25_daily_avgs[[i]] <- daily_pm25 
}

# Save into df 
pm25_daily_avgs <- cbind.data.frame(trt, pm25_daily_avgs)

# Setup data in long format 
pm25_long_fun <- function(df, values_to_label){
  pm25_daily_long <- df %>% 
    pivot_longer(cols = c(2:93), names_to = "Date", values_to = values_to_label) %>% 
    mutate(Date = as.Date(factor(Date, labels = seq.Date(as.Date("2023-05-01"), as.Date("2023-07-31"), by = "day"))))
  return(pm25_daily_long)
}

# Inner join with community areas 
pm25_daily_avgs_long <- pm25_long_fun(pm25_daily_avgs, "PM25_Avg") %>% 
  inner_join(trt_community_areas, by = "geoid") %>% 
  mutate(Neighborhood = factor(Neighborhood, 
                               levels = c("Southwest Side", "Far North Side", "Northwest Side", "North Side", "West Side", "Central",
                                          "South Side", "Far Southwest Side", "Far Southeast Side")))

```

```{r community area map, fig.width=10}

# Plot community area map 
community_areas %>% 
  mutate(Neighborhood = factor(Neighborhood, 
                               levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "South Side",
                                          "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>%
  ggplot() + 
  geom_sf(aes(fill = Neighborhood), name = "Region", linewidth = 0.2, color = "black") +
  scale_fill_brewer(palette = "Paired", name = "Region") +
  theme_bw() +
  theme(axis.line = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          #legend.position = "bottom",
          #legend.direction = "vertical",
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14, face = "bold"),
          plot.title = element_text(size = 16, face = "bold", hjust = 0.5)) + 
  ggtitle(label = "City of Chicago: Community Areas and Region Definitions")

```

# Assessing sociodemographic composition of neighborhood areas

```{r assessing sociodemographics of community areas and their respective neighborhoods}

neighborhood_summary <- community_areas_sociodemo %>% 
  group_by(Neighborhood) %>% 
  summarise(Population = sum(Population),
            pctNHwht = paste0(round(median(pctNHwht),digits = 1), " (", round(quantile(pctNHwht, 0.25), digits = 1), ", ", round(quantile(pctNHwht, 0.75), digits = 1), ")"),
            pctNHblk = paste0(round(median(pctNHblk),digits = 1), " (", round(quantile(pctNHblk, 0.25), digits = 1), ", ", round(quantile(pctNHblk, 0.75), digits = 1), ")"),
            pctHisp = paste0(round(median(pctHisp),digits = 1), " (", round(quantile(pctHisp, 0.25), digits = 1), ", ", round(quantile(pctHisp, 0.75), digits = 1), ")"),
            Unemployment_rate = paste0(round(median(Unemployment_rate),digits = 1), " (", round(quantile(Unemployment_rate,0.25), digits = 1), ", ", round(quantile(Unemployment_rate, 0.75), digits = 1), ")"),
            MedHHinc = paste0(round(median(MedHHinc, na.rm = T),digits = 1), " (", round(quantile(MedHHinc,0.25, na.rm = T), digits = 1), ", ", round(quantile(MedHHinc, 0.75, na.rm = T), digits = 1), ")"),
            Poverty_rate = paste0(round(median(Poverty_rate),digits = 1), " (", round(quantile(Poverty_rate,0.25), digits = 1), ", ", round(quantile(Poverty_rate, 0.75), digits = 1), ")"), 
            HS_grad_rate = paste0(round(median(HS_grad_rate),digits = 1), " (", round(quantile(HS_grad_rate,0.25), digits = 1), ", ", round(quantile(HS_grad_rate, 0.75), digits = 1), ")"),
            Coll_grad_rate = paste0(round(median(Coll_grad_rate),digits = 1), " (", round(quantile(Coll_grad_rate,0.25), digits = 1), ", ", round(quantile(Coll_grad_rate, 0.75), digits = 1), ")"),
            Uninsured_rate = paste0(round(median(Uninsured_rate),digits = 1), " (", round(quantile(Uninsured_rate,0.25), digits = 1), ", ", round(quantile(Uninsured_rate, 0.75), digits = 1), ")"),
            COPD_rate = paste0(round(median(COPD_rate),digits = 1), " (", round(quantile(COPD_rate,0.25), digits = 1), ", ", round(quantile(COPD_rate, 0.75), digits = 1), ")"),
            Disability_rate = paste0(round(median(Disability_rate),digits = 1), " (", round(quantile(Disability_rate,0.25), digits = 1), ", ", round(quantile(Disability_rate, 0.75), digits = 1), ")")) %>% 
  ungroup() %>% 
  rename("% NHW" = "pctNHwht",
         "% NHB" = "pctNHblk",
         "% Hispanic" = "pctHisp",
         "% Unemployed" = "Unemployment_rate",
         "% Poverty" = "Poverty_rate",
         "HS grad rate" = "HS_grad_rate", 
         "College grad rate" = "Coll_grad_rate",
         "% Uninsured" = "Uninsured_rate",
         "% COPD" = "COPD_rate",
         "% Disability" = "Disability_rate")

neighborhood_summary2 <- community_areas_sociodemo2 %>% 
  mutate(Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "South Side",
                                          "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  group_by(Neighborhood) %>% 
  summarise(Medicaid_rate = paste0(round(median(Medicaid_rate, na.rm = T),digits = 1), " (", round(quantile(Medicaid_rate, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Medicaid_rate, 0.75, na.rm = T), digits = 1), ")"),
            Major_crimes = paste0(round(median(Major_crimes, na.rm = T),digits = 1), " (", round(quantile(Major_crimes, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Major_crimes, 0.75, na.rm = T), digits = 1), ")"),
            Eviction_rate = paste0(round(median(Eviction_rate, na.rm = T),digits = 1), " (", round(quantile(Eviction_rate, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Eviction_rate, 0.75, na.rm = T), digits = 1), ")"),
            Walkability_index = paste0(round(median(Walkability_index),digits = 1), " (", round(quantile(Walkability_index,0.25), digits = 1), ", ", round(quantile(Walkability_index, 0.75), digits = 1), ")"),
            Rent_burdened_rate = paste0(round(median(Rent_burdened_rate, na.rm = T),digits = 1), " (", round(quantile(Rent_burdened_rate,0.25, na.rm = T), digits = 1), ", ", round(quantile(Rent_burdened_rate, 0.75, na.rm = T), digits = 1), ")"),
            Ever_had_cancer_rate = paste0(round(median(Ever_had_cancer_rate, na.rm = T),digits = 1), " (", round(quantile(Ever_had_cancer_rate,0.25, na.rm = T), digits = 1), ", ", round(quantile(Ever_had_cancer_rate, 0.75, na.rm = T), digits = 1), ")"), 
            Crowded_housing_rate = paste0(round(median(Crowded_housing_rate, na.rm = T),digits = 1), " (", round(quantile(Crowded_housing_rate,0.25, na.rm = T), digits = 1), ", ", round(quantile(Crowded_housing_rate, 0.75, na.rm = T), digits = 1), ")")) %>% 
  ungroup() 

neighborhood_summary3 <- community_areas_sociodemo3 %>% 
  group_by(Neighborhood) %>% 
  summarise(Obesity_rate = paste0(round(median(Obesity_rate, na.rm = T),digits = 1), " (", round(quantile(Obesity_rate, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Obesity_rate, 0.75, na.rm = T), digits = 1), ")"),
            Hypertension_rate = paste0(round(median(Hypertension_rate, na.rm = T),digits = 1), " (", round(quantile(Hypertension_rate, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Hypertension_rate, 0.75, na.rm = T), digits = 1), ")"),
            Diabetes_rate = paste0(round(median(Diabetes_rate, na.rm = T),digits = 1), " (", round(quantile(Diabetes_rate, 0.25, na.rm = T), digits = 1), ", ", round(quantile(Diabetes_rate, 0.75, na.rm = T), digits = 1), ")"),
            Asthma_rate = paste0(round(median(Asthma_rate, na.rm = T),digits = 1), " (", round(quantile(Asthma_rate,0.25, na.rm = T), digits = 1), ", ", round(quantile(Asthma_rate, 0.75,  na.rm = T), digits = 1), ")"),
            Cancer_diagnosis_rate = paste0(round(median(Cancer_diagnosis_rate, na.rm = T),digits = 1), " (", round(quantile(Cancer_diagnosis_rate,0.25, na.rm = T), digits = 1), ", ", round(quantile(Cancer_diagnosis_rate, 0.75, na.rm = T), digits = 1), ")"),
            Life_expectancy = paste0(round(median(Life_expectancy, na.rm = T),digits = 1), " (", round(quantile(Life_expectancy,0.25, na.rm = T), digits = 1), ", ", round(quantile(Life_expectancy, 0.75, na.rm = T), digits = 1), ")")) %>% 
  ungroup() 

neighborhood_summary4 <- community_areas_sociodemo4 %>% 
  mutate(Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "South Side",
                                          "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  group_by(Neighborhood) %>% 
  summarise(pct_0_4 = paste0(round(median(pct_0_4, na.rm = T),digits = 1), " (", round(quantile(pct_0_4, 0.25, na.rm = T), digits = 1), ", ", round(quantile(pct_0_4, 0.75, na.rm = T), digits = 1), ")"),
             pct_0_17 = paste0(round(median(pct_0_17, na.rm = T),digits = 1), " (", round(quantile(pct_0_17, 0.25, na.rm = T), digits = 1), ", ", round(quantile(pct_0_17, 0.75, na.rm = T), digits = 1), ")"),
             pct_65p = paste0(round(median(pct_65p, na.rm = T),digits = 1), " (", round(quantile(pct_65p, 0.25, na.rm = T), digits = 1), ", ", round(quantile(pct_65p, 0.75, na.rm = T), digits = 1), ")")) %>% 
  ungroup()

neighborhood_summary5 <- community_areas_sociodemo5 %>% 
  mutate(Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "South Side",
                                          "Southwest Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  group_by(Neighborhood) %>% 
  summarise(ADMR = paste0(round(median(ADMR, na.rm = T),digits = 1), " (", round(quantile(ADMR, 0.25, na.rm = T), digits = 1), ", ", round(quantile(ADMR, 0.75, na.rm = T), digits = 1), ")")) %>% 
  ungroup()

neighborhood_summary %>% kbl(caption = "Table 1. Mean (SD) of Sociodemographic Characteristics in Neighborhood Areas") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

neighborhood_summary2 %>% kbl(caption = "Table 1 Continued") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

neighborhood_summary3 %>% kbl(caption = "Table 1 Continued") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

neighborhood_summary4 %>% kbl(caption = "Table 1 Continued") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

neighborhood_summary5 %>% kbl(caption = "Table 1 Continued") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

```

# Distribution of Daily PM2.5 concentrations 

```{r daily pm25 concentrations, fig.width=10}

pm25_daily_avgs_long %>% 
  ggplot() + 
  geom_boxplot(aes(x = Date, y = PM25_Avg, group = Date), outlier.shape = NA) + 
  scale_x_date(breaks = as.Date(c("2023-05-01", "2023-05-15", "2023-05-31", "2023-06-15", "2023-06-27", "2023-06-28", "2023-07-15", "2023-07-31")), 
               labels = c("5/1", "5/15", "5/31", "6/15", "6/27-28", "", "7/15", "7/31")) +
  geom_vline(aes(xintercept = as.Date("2023-06-27")), linetype = 2, size = 1.1, color = "firebrick") + 
  geom_hline(aes(yintercept = 35), linetype = 2, size = 1.1, color = "green3") + 
  geom_hline(aes(yintercept = 15.3), linetype = 2, size = 1.1, color = "steelblue") + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16),
        panel.grid = element_blank(),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20),
        plot.title = element_text(size = 18)) +
     xlab("Date") + ylab("Daily Average PM2.5") #+ ggtitle(label = "Distribution of daily average tract-level PM2.5 concentrations", subtitle = "Vertical line represents the start of the peak WFS period.")

```

# Examining PM2.5 concentrations in neighborhoods at different points during the study period  

## Daily comparisons 

The table below shows daily average PM2.5 concentrations by neighborhood, and evaluates significant differences between them via a one-way ANOVA. The daily p-value associated with the one-way ANOVA is included in the table. 

```{r assessing PM2.5 concentrations for community areas and their respective neighborhoods at different points during the study period - daily comparisons}

# For each day, calculate mean and sd of pm2.5 exposure for each of the neighborhoods, and capture sig differences between neighborhoods 
neighborhood_daily_avgs <- list()
anova_pvals <- vector()
for (i in unique(as.character(pm25_daily_avgs_long$Date))) {
  pm25_day <- pm25_daily_avgs_long %>% filter(Date == as.Date(i))
  neighborhood_daily_avgs[[i]] <- pm25_day %>% 
    group_by(Neighborhood) %>% 
    summarise(`PM2.5 Mean (SD)`= paste0(round(mean(PM25_Avg),1), " (", round(sd(PM25_Avg),1), ")")) %>% 
    pivot_wider(names_from = Neighborhood, values_from = `PM2.5 Mean (SD)`)
  
  anova_pvals[i] <- round(summary(aov(PM25_Avg ~ Neighborhood, data = pm25_daily_avgs_long  %>% filter(Date == as.Date(i))))[[1]]$`Pr(>F)`[1],3)
  
}

bind_rows(neighborhood_daily_avgs) %>% 
  mutate(Date = format(seq.Date(from = as.Date("2023-05-01"), to = as.Date("2023-07-31"), by = "day"), "%m/%d"),
        `p-value` = anova_pvals) %>% 
  select(10, 1:9, 11) %>% 
  kbl(caption = "Daily PM2.5 Mean (SD) Exposure by Neighborhood") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

```

## Period averages 

* Below we highlight average PM2.5 concentrations at the neighborhood-level across the following time periods:
    + Pre-peak WF days (May 1 - June 26)
    + Peak WF days (June 27 - June 28)
    + Post-peak WF days (June 29 - July 31)
    
```{r assessing PM2.5 concentrations for community areas and their respective neighborhoods at different points during the study period - period averages, fig.width=10}

pm25_period_avgs <- pm25_daily_avgs_long %>% 
  mutate(Period = factor(if_else(Date <= as.Date("2023-06-26"), "Pre-peak",
                          if_else(Date >= as.Date("2023-06-27") & Date <= as.Date("2023-06-28"), "Peak", "Post-peak")),
                         levels = c("Pre-peak", "Peak", "Post-peak")))

# For each designated period, calculate mean and sd of pm2.5 exposure for each of the neighborhoods, and capture sig differences between neighborhoods 
neighborhood_period_avgs <- list()
anova_pvals <- vector()
for (i in unique(as.character(pm25_period_avgs$Period))) {
  pm25_period <- pm25_period_avgs %>% filter(Period == i)
  neighborhood_period_avgs[[i]] <- pm25_period %>% 
    group_by(Neighborhood) %>% 
    summarise(`PM2.5 Mean (SD)`= paste0(round(mean(PM25_Avg),1), " (", round(sd(PM25_Avg),1), ")")) %>% 
    pivot_wider(names_from = Neighborhood, values_from = `PM2.5 Mean (SD)`)
  
  anova_pvals[i] <- round(summary(aov(PM25_Avg ~ Neighborhood, data = pm25_period_avgs  %>% filter(Period == i)))[[1]]$`Pr(>F)`[1],3)
  
}

bind_rows(neighborhood_period_avgs) %>% 
  mutate(Period = c("Pre-peak", "Peak", "Post-peak"),
         `p-value` = anova_pvals) %>% 
  select(10, 1:9, 11) %>% 
  kbl(caption = "PM2.5 Mean (SD) Exposure by Neighborhood for Designated Time Periods") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F, position = "center", fixed_thead  = T) 

```

### Spatial distribution 

```{r assessing PM2.5 concentrations for community areas and their respective neighborhoods at different points during the study period - period averages - spatial dist, eval=FALSE}

# Join pm2.5 period averages with tract boundaries 
pm25_period_avgs_sf <- pm25_period_avgs %>%
  group_by(geoid, Period) %>%
  mutate(PM25_Avg = mean(PM25_Avg)) %>% 
  distinct(geoid, .keep_all = T) %>%
  ungroup() %>% 
  mutate(Period2 = if_else(Period == "Peak", "Peak", "Non-Peak")) %>% 
  inner_join(trt_boundaries, by = c("geoid", "Neighborhood")) %>% 
  select(-Community.x, -Community.y) %>% 
  group_by(Period2) %>% 
  mutate(PM25_Avg = cut(PM25_Avg, quantile(PM25_Avg, seq(0,1,0.2)), include.lowest = T)) %>% 
  ungroup() %>% 
  mutate(PM25_Avg2 = str_replace_all(gsub("\\[|\\]|\\(|\\)", "", PM25_Avg), pattern = ",", replacement = " - ")) %>% 
  st_as_sf()

factor_levels <- pm25_period_avgs_sf %>% 
  st_set_geometry(NULL) %>% 
  group_by(PM25_Avg, PM25_Avg2) %>% 
  distinct(PM25_Avg, .keep_all = F) %>% 
  ungroup() %>% 
  arrange(PM25_Avg)

pm25_period_avgs_sf$PM25_Avg <- factor(pm25_period_avgs_sf$PM25_Avg, labels = factor_levels$PM25_Avg2)

# Create map function 
map_fun <- function(period, viridis_option){
  period_file_name <- ifelse(period == "Pre-peak", "Pre-peak",
                             ifelse(period == "Peak", "Peak", "Post-peak"))
  
  ggplot() + 
    geom_sf(aes(fill = PM25_Avg), linewidth = 0.15, color = "black", data = pm25_period_avgs_sf %>% filter(Period == period)) + 
    scale_fill_viridis_d(direction = -1, option = viridis_option, name = "PM2.5") +
    theme_bw() +
    theme(axis.line = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          #legend.position = "bottom",
          #legend.direction = "vertical",
          legend.text = element_text(size = 12),
          legend.title = element_text(size = 14, face = "bold"),
          plot.title = element_text(size = 16, face = "bold", hjust = 0.5)) + 
     ggtitle(label = period)
}

map_fun("Pre-peak", "inferno")
map_fun("Peak", "inferno")
map_fun("Post-peak", "inferno")

```

# Modeling PM2.5 exposure using an ITS approach

We model PM2.5 exposure using an ITS modeling approach to better compare exposure levels before, during, and directly after the start of the peak WF smoke.  

* In its simplest form, the ITS model is specified as follows: 
    + **Y = Beta0 + Beta1xT + Beta2xD + Beta3xP + e**, where:
        + Beta0 is the baseline level of the outcome
        + T = time variable passed from the start of the observation period
            + Beta1 = slope of the line before the intervention 
        + D = treatment variable where 1 is after the intervention and 0 is before the intervention
            + Beta2 = **immediate effect** after the intervention 
        + P = time variable indicating time passed since the intervention (before the intervention, the value is set to 0)
            + Beta3 = **sustained effect** (i.e., difference between the slope of the line prior to the intervention and the slope of the line subsequent to the intervention)

* For the purpose of this study, we define each of the three time covariates as follows:  
    + Y = PM2.5 daily average 
    + T = 1, 2, 3,..., 92 (represents sequential days from 5/1-7/31)
    + D = 0 (all days before 6/27), 1 (all days after and including 6/27)
        + We define the intervention as the first peak WF day (6/27)
    + P = 0 (all days before 6/29); 1, 2, 3,..., 33 (represents sequential days from 6/29-7/31)
    
* Further model specifications:
    + For all set of models, we include a random effect (random intercept) component for census tracts 
    + To deal with likely time series related autocorrelation, we include a first-order autoregressive correlation structure in the model to allow for correlation of time 

```{r modeling PM2.5 exposure using an interrupted time series approach - setup data for modeling}

# Set-up df for ITS 
pm25_its_df <- pm25_daily_avgs_long

pm25_its_df$Time <- as.numeric(as.factor(pm25_its_df$Date))
pm25_its_df$Intervention <- as.factor(if_else(pm25_its_df$Date < as.Date("2023-06-27"), 0, 1))
pm25_its_df$Post_Intervention_Time <- rep(c(rep(0,59), 1:33), length(unique(pm25_its_df$geoid)))

```

## ITS model via segmented linear regression

Our base ITS model takes the form of a **segmented linear mixed-effects** model. 

Below are the model results, and a plot of the fitted values. 

```{r modeling PM2.5 exposure using an interrupted time series approach - fit both sets of and save output model, eval=FALSE}

base_mod <- lme(PM25_Avg ~ Time + Intervention + Post_Intervention_Time,
                random = ~1|geoid,
    correlation = corARMA(p = 1, q = 2, form = ~Time), 
    data = pm25_its_df, method = "ML")

base_mod2 <- lme(PM25_Avg ~ Time*Neighborhood + Intervention*Neighborhood + Post_Intervention_Time*Neighborhood,
                random = ~1|geoid,
    correlation = corARMA(p = 1, q = 2, form = ~Time), 
    data = pm25_its_df, method = "ML")

saveRDS(base_mod, "~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/base_mod_v2.rds")
saveRDS(base_mod2, "~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/base_mod2_v3.rds")

```

```{r modeling PM2.5 exposure using an interrupted time series approach - base model 1 results}

base_mod <- readRDS("~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/base_mod_v2.rds")
base_mod2 <- readRDS("~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/base_mod2_v3.rds")
sjPlot::tab_model(base_mod, dv.labels = "PM2.5 Avg")

```

* **Interpretation of model results:**     
    + Prior to the intervention (i.e., the first peak WF day), the average PM2.5 concentration was increasing by 0.06 units per day
    + At the first peak WF day, there was a positive step change of 37.67, meaning that the average increase in PM2.5 for census tracts from 6/26-6/27 was ~38 units
    + After the first peak WF day, average PM2.5 concentrations began decreasing -1.56 units per day 

```{r modeling PM2.5 exposure using an interrupted time series approach - base model figure}

pm25_its_df %>% 
  mutate(model_predictions = predictSE.lme(base_mod, ., se.fit=T)$fit,
         model_se = predictSE.lme(base_mod, ., se.fit = T)$se) %>%
  ggplot(aes(x = Time, PM25_Avg)) +
  geom_ribbon(aes(ymin = model_predictions - (1.96*model_se),
                  ymax = model_predictions + (1.96*model_se)), fill = "lightblue") + 
  geom_line(aes(x  = Time, y = model_predictions), size = 0.8, color = "black") +
  geom_vline(aes(xintercept = 58), linetype = 2) + 
  geom_point(alpha = 0.1, color = "steelblue4") +
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        plot.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12, face = "bold")) + 
  xlab("Time") + 
  ylab("PM2.5 Avg") + 
  ggtitle(label = "Base model: Observations vs. predicted values",
          subtitle = "Blue points = Actual PM2.5 concentrations\nBlack line = Model fitted values\nDashed line = Start of the WF period (intervention)")

```

### Model with neighborhood interaction 

To assess differences between neighborhoods in these associations, we interact each of the time covariates in the ITS model with neighborhood, using the Southwest Side as reference. 

```{r modeling PM2.5 exposure using an interrupted time series approach - int term for neighborhood}

sjPlot::tab_model(base_mod2, dv.labels = "PM2.5 Avg")

time_coef <- paste0("Time:Neighborhood", levels(pm25_its_df$Neighborhood))
intervention_coef <- paste0("Neighborhood", levels(pm25_its_df$Neighborhood), ":Intervention1")
post_intervention_coef <- paste0("Neighborhood", levels(pm25_its_df$Neighborhood), ":Post_Intervention_Time")


# Pre-peak WFS plot 
sjPlot::plot_model(base_mod2, type = "est", terms = time_coef) + 
  scale_color_manual(values = c("blue", "red")) + 
  geom_hline(aes(yintercept = 0), linetype = "dashed", color = "gray50") +
  scale_x_discrete(labels = paste0("Pre-peak WFS:", rev(levels(pm25_its_df$Neighborhood)[-1]))) +
  theme_bw() + 
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14),
        plot.title = element_text(size = 16)) + 
  ylab("Coefficient Estimates") +
  ggtitle(label = "Pre-peak WFS x Region") 

# Peak WFS plot 
sjPlot::plot_model(base_mod2, type = "est", terms = intervention_coef) + 
  scale_color_manual(values = c("blue", "red")) + 
  geom_hline(aes(yintercept = 0), linetype = "dashed", color = "gray50") +
  scale_x_discrete(labels = paste0("Peak WFS:", rev(levels(pm25_its_df$Neighborhood)[-1]))) +
  theme_bw() + 
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_text(size = 17),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 22)) + 
  ggtitle(label = "Peak WFS x Region") 

# Post-peak WFS plot 
sjPlot::plot_model(base_mod2, type = "est", terms = post_intervention_coef) + 
  scale_color_manual(values = c("blue", "red")) + 
  geom_hline(aes(yintercept = 0), linetype = "dashed", color = "gray50") +
  scale_x_discrete(labels = paste0("Post-peak WFS:", rev(levels(pm25_its_df$Neighborhood)[-1]))) +
  theme_bw() + 
  theme(
        panel.grid = element_blank(),
        axis.title = element_text(size = 24),
        axis.text = element_text(size = 17),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 22)) + 
  ylab("Coefficient Estimates") +
  ggtitle(label = "Post-peak WFS x Region") 

```
  
## ITS model using GAM

From our observed data, we know the time trend in the study is non-linear. Below, we use a GAM to accommodate the nonlinear trend in PM2.5 exposure. For each of the continuous time covariates - T and P - we apply thin plate regression splines (default smoothing function in R) with 4 and 5 knots, respectively. We choose 4 and 5 knots (default is 9), to avoid overfitting while still capturing the non-linear trend in the data. 

Below are the results of the model and a plot of the fitted values. 

```{r modeling PM2.5 exposure using an interrupted time series approach - fit and save gam models, eval=FALSE}

# Fit basic gam model
gam_mod <- gamm(PM25_Avg ~ s(Time, k = 4) + Intervention + s(Post_Intervention_Time, k = 5),
                random = list(geoid=~1),
                correlation = corARMA(p = 1, q = 2, form = ~Time),
                data = pm25_its_df, method = "ML")

# Fit gam model with different smoothing parameters for each level of region (neighborhood)
gam_mod2 <- gamm(PM25_Avg ~ s(Time, by = Neighborhood, k = 4) + Intervention*Neighborhood + s(Post_Intervention_Time, by = Neighborhood, k = 5),
                random = list(geoid=~1),
                correlation = corARMA(p = 1, q = 2, form = ~Time),
                data = pm25_its_df, method = "ML")
gam_mod3 <- gamm(PM25_Avg ~ s(Time, k = 4) + Intervention + s(Post_Intervention_Time, by = Neighborhood, k = 5),
                random = list(geoid=~1),
                correlation = corARMA(p = 1, q = 2, form = ~Time),
                data = pm25_its_df, method = "ML")

# Save models 
saveRDS(gam_mod, "~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod1_v8.rds")
saveRDS(gam_mod2, "~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod2_v9.rds")
saveRDS(gam_mod3, "~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod3_v2.rds")

```

```{r modeling PM2.5 exposure using an interrupted time series approach - display model results and plot, fig.width=10}

gam_mod <- readRDS("~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod1_v8.rds")
gam_mod2 <- readRDS("~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod2_v9.rds") 
gam_mod3 <- readRDS("~/Documents/CEHI/Chicago_Wildfire/Data/Output/Models/gam_mod3_v2.rds") 

pm25_its_df %>% 
  mutate(model_predictions = predict.gam(gam_mod$gam, ., se.fit=T)$fit,
         model_se = predict.gam(gam_mod$gam, ., se.fit = T)$se) %>%
  ggplot(aes(x = Date, PM25_Avg)) +
  geom_ribbon(aes(ymin = model_predictions - (1.96*model_se),
                  ymax = model_predictions + (1.96*model_se)), fill = "lightblue") + 
  geom_line(aes(x  = Date, y = model_predictions), size = 0.8, color = "black") +
  scale_x_date(breaks = as.Date(c("2023-05-01", "2023-05-15", "2023-05-31", "2023-06-15", "2023-06-27", "2023-07-15", "2023-07-31")), date_labels = "%m/%d") +
  geom_vline(aes(xintercept = as.Date("2023-06-27")), linetype = 2, size = 1, color = "forestgreen") + 
  geom_hline(aes(yintercept = 35), linetype = 4, size = 1, color = "firebrick4") + 
  geom_point(alpha = 0.1, color = "steelblue4") +
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        plot.title = element_text(size = 16, face = "bold"),
        axis.text = element_text(size = 18),
        axis.title = element_text(size = 18, face = "bold")) + 
  xlab("Time") + 
  ylab("PM2.5 Average") + 
  ggtitle(label = "GAM: Observations vs. predicted values")

```

### ITS model using GAM: Allowing smoothing parameters to vary between neighborhoods 

Using GAM, we can allow the smoothing parameters to vary between neighborhoods, allowing us to compare PM2.5 exposure levels between neighborhoods at different points in time. 

Below are the model results and the plot of the fitted values. 

```{r modeling PM2.5 exposure using an interrupted time series approach - allowing smoothing parameters to vary between neighborhoods - display model results and plot, fig.width=12}

summary(gam_mod2$gam)

pm25_its_df %>% 
  mutate(model_predictions = predict.gam(gam_mod2$gam, ., se.fit=T)$fit,
         model_se = predict.gam(gam_mod2$gam, ., se.fit = T)$se) %>%
  ggplot(aes(x = Date, PM25_Avg)) +
  geom_ribbon(aes(ymin = model_predictions - (1.96*model_se),
                  ymax = model_predictions + (1.96*model_se)), fill = "lightblue") + 
  geom_line(aes(x  = Date, y = model_predictions, color = Neighborhood), size = 1.1) +
  scale_x_date(breaks = as.Date(c("2023-05-01", "2023-05-15", "2023-05-31", "2023-06-15", "2023-06-27", "2023-07-15", "2023-07-31")), date_labels = "%m/%d") + 
  geom_vline(aes(xintercept = as.Date("2023-06-27")), linetype = 2, size = 1.2, color = "forestgreen") + 
  geom_hline(aes(yintercept = 35), linetype = 4, size = 1.2, color = "firebrick4") + 
  geom_point(alpha = 0.1, color = "steelblue4") +
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.text = element_text(size = 17),
        axis.title = element_text(size = 22, face = "bold"),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 22)) + 
  labs(color = "Region") +
  xlab("Date") + 
  ylab("Daily Average PM2.5")


# Plot intervention x Neighborhood interaction 
model_coefficients <- summary(gam_mod2$gam)[["p.coeff"]][11:18]
model_standard_errors <- summary(gam_mod2$gam)[["se"]][11:18]

data.frame(Region = factor(levels(pm25_its_df$Neighborhood)[-1], levels = levels(pm25_its_df$Neighborhood)[-1]),
           Coefficient = unname(model_coefficients),
           SE = unname(model_standard_errors),
           Lower_CI = unname(model_coefficients)  - 1.96*unname(model_standard_errors),
           Upper_CI = unname(model_coefficients)  + 1.96*unname(model_standard_errors)) %>% 
  ggplot(aes(x = Region, y = Coefficient, color = Region, group = Region)) + 
  geom_point(size = 4) + 
  geom_errorbar(aes(ymin=Lower_CI, ymax=Upper_CI), width = 0.4, size = 1.3) + 
  geom_hline(aes(yintercept = 0), linetype = 2, size = 2) + 
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(-8.2, 6), breaks = seq(-8,5,2)) + 
  annotate("text", x = 1.4, y = 6, label = "- - - Southwest Side", 
           color = "black", size = 10) + 
  theme_bw() + 
  theme(#panel.grid = element_blank(),
        legend.position = "none",
        axis.text.x = element_text(size = 26, angle = 20, vjust = 0.65),
        axis.text.y = element_text(size = 26),
        axis.title.y = element_blank(),
        axis.title.x = element_text(size = 28))# +
  #ylab("Difference in PM2.5") 

```

#### Assessing pairwise differences between neighborhood smooth interactions 

```{r modeling PM2.5 exposure using an interrupted time series approach - allowing smoothing parameters to vary between neighborhoods - assessing pairwise differences between neighborhood smooth interactions - setup and display data, fig.width=12, fig.height=8}

# Calculate difference in smooths between pre-peak and post-peak 
sm_dif_time <- difference_smooths(gam_mod2, select = "s(Time)", group_means = T) %>% filter(Time <= 57) 
sm_dif_postint <- difference_smooths(gam_mod3, select = "s(Post_Intervention_Time)", group_means = T) %>%
  filter(Post_Intervention_Time >= 1 & Post_Intervention_Time <= 33)

  
sm_dif_time_opposite <- sm_dif_time %>% 
  select(1:2,4,3,5:6,8,7,9) %>% 
  rename(.level_1 = .level_2,
         .level_2 = .level_1,
         .upper_ci = .lower_ci,
         .lower_ci = .upper_ci) %>% 
  mutate(.diff = .diff*-1,
         .lower_ci = .lower_ci*-1,
         .upper_ci = .upper_ci*-1)

sm_dif_postint_opposite <- sm_dif_postint %>% 
  select(1:2,4,3,5:6,8,7,9) %>% 
  rename(.level_1 = .level_2,
         .level_2 = .level_1,
         .upper_ci = .lower_ci,
         .lower_ci = .upper_ci) %>% 
  mutate(.diff = .diff*-1,
         .lower_ci = .lower_ci*-1,
         .upper_ci = .upper_ci*-1)

sm_dif_fun <- function(time_covariate, neighborhood){
  sm_dif_objs <- if(time_covariate == "Time"){
     list(sm_dif_time, sm_dif_time_opposite)
   } else {
     list(sm_dif_postint, sm_dif_postint_opposite)
   }
  
  sm_dif_neighborhood <- sm_dif_objs[[1]] %>% filter(.level_1 == neighborhood)
  # Save neighborhood comparisons not captured 
  all_neighborhoods <- as.character(unique(pm25_daily_avgs_long$Neighborhood))
  missing_comparisons <- all_neighborhoods[!all_neighborhoods %in% c(unique(sm_dif_neighborhood$.level_2), neighborhood)]
  # For those missing comparisons, index the appropriate opposite df created above 
  comparisons_to_add <- sm_dif_objs[[2]] %>% 
    filter(.level_1 == neighborhood & .level_2 %in% c(missing_comparisons))
  
  # Bind rows to capture all comparisons in 1 df 
  bind_rows(sm_dif_neighborhood, comparisons_to_add)
}

# Run function across all neighborhoods, and for both time covariates included in the GAM 
sm_dif_time_list <- list()
sm_dif_postint_list <- list()
for (i in as.character(unique(pm25_daily_avgs_long$Neighborhood))) {
  sm_dif_time_list[[i]] <- sm_dif_fun("Time", i)
  sm_dif_postint_list[[i]] <- sm_dif_fun("Post_Intervention_Time", i)
}

# Bind rows 
sm_dif_time_df <- bind_rows(sm_dif_time_list)
sm_dif_postint_df <- bind_rows(sm_dif_postint_list) %>% rename(Time = Post_Intervention_Time)

# Combine both pre-peak and post-peak dfs 
sm_dif <- bind_rows(sm_dif_time_df, sm_dif_postint_df) %>% 
  rename("Period" = ".smooth") %>% 
  mutate(Period = factor(if_else(Period == "s(Time)", "Pre-peak", "Post-peak"), levels = c("Pre-peak", "Post-peak"))) 

# Create pairwise difference plot function to output plots more easily 
pairwise_diff_plot_fun <- function(ref_neighborhood){
  sm_dif %>% 
    mutate(.level_1 = factor(.level_1, levels = levels(pm25_daily_avgs_long$Neighborhood))) %>% 
    filter(.level_2 == ref_neighborhood) %>% 
    ggplot(aes(x = Time, y = .diff, color = Period)) + 
    geom_line() +
    geom_ribbon(aes(ymin = .lower_ci, ymax = .upper_ci, fill = Period), alpha=0.1) + 
    geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
    scale_color_brewer(palette = "Dark2") +
    scale_fill_brewer(palette = "Dark2") +
    scale_x_continuous(breaks = seq(0,57,10)) +
    theme_bw() + 
    theme(legend.position = "bottom",
          legend.text = element_text(size = 18),
          legend.title = element_text(size = 18, face = "bold"),
          panel.grid = element_blank(),
          axis.title = element_text(size = 18, face = "bold"),
          axis.text = element_text(size = 18),
          plot.caption = element_text(hjust = 0, size = 12),
          strip.text = element_text(size = 18, face = "bold")) +
    facet_wrap(~.level_1) + 
    ylab("Difference") + 
    xlab("Time, days") + 
    labs(caption =  paste0("Horizontal vertical line = ", ref_neighborhood, "\nPre-peak period = May 1 - June 26 (57 days)\nPost-peak period = June 29 - July 31 (33 days)"))
}

# Output plots for Southwest Side and Far Southwest Side
pairwise_diff_plot_fun("Southwest Side")
pairwise_diff_plot_fun("Far Southwest Side")

```

# Figure showing how many days it took to return to baseline by region 

```{r figure showing how many days it took to return to baseline by region, fig.width=10}

# First, let's calculate overall average pre-peak WFS PM2.5 concentrations 
pre_peak_overall_avg <- as.numeric(pm25_its_df %>% 
  filter(Date <= as.Date("2023-06-26")) %>% 
  summarise(PM25_Avg = mean(PM25_Avg)))


return_to_baseline <- pm25_its_df %>% 
  filter(Date >= as.Date("2023-06-27")) %>% 
  group_by(Neighborhood, Date) %>%
  summarise(PM25_Avg = mean(PM25_Avg)) %>% 
  ungroup() %>% 
  group_by(Neighborhood) %>% 
  mutate(PM25_3day_MA = rollmean(PM25_Avg, k = 3, fill = NA, align = "right")) %>% 
  ungroup() %>% 
  drop_na() %>% 
  filter(Date <= as.Date("2023-07-08")) %>% 
  mutate(vline_pos = if_else(Neighborhood == "Southwest Side", as.Date("2023-07-07"),
                             if_else(Neighborhood == "Far North Side", as.Date("2023-07-03"),
                                     if_else(Neighborhood == "Northwest Side", as.Date("2023-07-03"), 
                                             if_else(Neighborhood == "North Side", as.Date("2023-07-03"),
                                                     if_else(Neighborhood == "West Side", as.Date("2023-07-03"),
                                                             if_else(Neighborhood == "Central", as.Date("2023-07-07"),
                                                                     if_else(Neighborhood == "South Side", as.Date("2023-07-07"),
                                                                             if_else(Neighborhood == "Far Southwest Side", as.Date("2023-07-08"), as.Date("2023-07-07"))))))))))

return_to_baseline %>% 
  mutate(Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "Southwest Side", "South Side", "Far Southwest Side", "Far Southeast Side"))) %>% 
  ggplot(aes(x = Date, y = PM25_3day_MA)) + 
  geom_point(size = 2) + 
  geom_line() + 
  geom_hline(aes(yintercept = pre_peak_overall_avg), size = 1.2, linetype = "dashed") + 
  facet_wrap(~Neighborhood) + 
  geom_vline(aes(xintercept = vline_pos), size = 1.2, color = "red", linetype = "dashed") + 
  scale_x_date(breaks = unique(return_to_baseline$Date), labels = as.character(3:12)) + 
  theme_bw() + 
  theme(panel.grid = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 14, face = "bold")) + 
  xlab("Days After the Start of the Peak WFS Period") + 
  ylab("PM2.5 3-Day Moving Average")

```

# Supplemental figure showing difference in PM2.5 concentrations by region with a reference value of 35 ug/m3

```{r supp figure, fig.width=10}

pm25_period_avgs %>% 
  group_by(Date, Period, Neighborhood) %>% 
  summarise(PM25_Avg = mean(PM25_Avg)) %>%
  ungroup() %>% 
  mutate(Neighborhood = factor(Neighborhood, levels = c("Far North Side", "Northwest Side", "North Side", "West Side", "Central", "Southwest Side", "South Side", "Far Southwest Side", "Far Southeast Side")),
         Difference = PM25_Avg - 35) %>% 
  ggplot(aes(x = Date, y = Difference)) + 
  geom_line(size = 1.2) + 
  scale_x_date(breaks = as.Date(c("2023-05-01", "2023-05-15", "2023-05-31", "2023-06-15", "2023-06-27", "2023-06-28", "2023-07-15", "2023-07-31")), 
               labels = c("5/1", "5/15", "5/31", "6/15", "6/27-28", "", "7/15", "7/31")) +
  geom_vline(aes(xintercept = as.Date("2023-06-27")), linetype = 2, size = 1.2, color = "firebrick") +
  geom_hline(aes(yintercept = 0), linetype = 2, size = 1.2, color = "steelblue3") +
  facet_wrap(~Neighborhood) + 
  theme_bw() + 
  theme(legend.title = element_text(size = 14, face = "bold"),
        panel.grid = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        axis.text = element_text(size = 12),
        strip.text = element_text(size = 14, face = "bold")) + 
  ylab("Difference in PM2.5")

```
